Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-05-05 issues.Rmd
commented out our notes and made small changes.

Below, we work through our issues we identified above, starting first with the unconventional use of a smoother when calculating the AUC-ROC scores. We reproduce Table 1 and Table 2 from @blair2020forecasting to show that their results are dependent on the unconventional use of a smoother.

We then discuss... (RM: I'll fill this out as I work through the evidence.)

@@

@@

```{r, message=FALSE}
auc <- read_csv("data/table1-redone.csv") %>%
  mutate(specification = factor(specification, levels = c("escalation", "quad", "goldstein", "CAMEO", "avg"))) %>%
  arrange(model, horizon, specification) %>% 
  pivot_longer(-c(model, horizon, specification), names_to = "ROC Smoothed", values_to = "Value") %>% 
  pivot_wider( names_from = "specification", values_from = "Value") %>% 
  mutate(`ROC Smoothed` = ifelse(`ROC Smoothed` == "smoothed", "Yes", "No"))

auc %>%
  knitr::kable(digits = 2, booktabs = TRUE, caption = "Comparison of smoothed and not smoothed AUC-ROC values") %>% 
  # column_spec(c(3), bold = TRUE) %>%
  collapse_rows(1:2, row_group_label_position = 'stack')

@@

- Improperly tuning of models
  - RM: In a twitter DM, I asked Blair to clarify -- "[I] was wondering how you all came to your tuning procedure for the RF model hyper-parameters?" He responded "mostly trial and error, honestly. ... Trial and error was on early data. Forecasts were for much later data"
- Using a regression framework in a classification problem
- Rounding of AUC/ROC scores
    + AB: smoothing or rounding? I think rounding is ok. 
- Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework
    + It's hacky but it works. The resulting forecasts are within the 0-1 interval.

@@

4. Design choices

Rick, I think the points below are more subjective choices. The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original Blair and Sambanis findings. The 3rd point about the RF models is somewhere between objective error and subjective choice. I'm not sure. But the points below I would say are subjective choices that could alter their findings and that I think are justifiable criticisms, but I think it'll be easier for someone to push back on this if they wanted to.

- Their train/test approach 
  - 5-year forecasts for sep plots and AUC scores)
- Smoothing of AUC/ROC scores
- Lack of yearly test forecasts in favor of a single 5-year test forecast
- The lack of procedures to account for rare events in an RF model

To-do

1. Full replication
    - Can we get their results
    - RM: maybe I can go through their code and bring it "up-to-date" i.e. make it tidy... 
        + AB: that sounds like a lot of work
    - Figure out how they built table 4 (It looks like this is done in STATA)
    - Check to see if it matter that they are using regression for a classification problem
        + AB: i tried switching the models as they have them just from regression to classification and it breaks because the data samples end up being all 0's. 
2. Address tuning 
    - Does properly tuning the CAMEO model improve its performance?
3. Address AUC/ROC smoothing... 
    - How drastic is the change; does it change their conclusions? 
4. Run a yearly test forecast -- as is the current standard in TSCS forecasts (Is it the current standard?) 
5.

>>>>

<!-- Below, we work through our issues we identified above, starting first with the unconventional use of a smoother when calculating the AUC-ROC scores. We reproduce Table 1 and Table 2 from @blair2020forecasting to show that their results are dependent on the unconventional use of a smoother.  -->

<!-- We then discuss... (RM: I'll fill this out as I work through the evidence.)  -->

@@

![](figures/figure1-replicated.png)

@@

```{r, message=FALSE}
auc <- read_csv("data/table1-redone.csv") %>%
  mutate(specification = factor(specification, levels = c("escalation", "quad", "goldstein", "CAMEO", "avg"))) %>%
  arrange(model, horizon, specification) %>% 
  pivot_longer(-c(model, horizon, specification), names_to = "ROC Smoothed", values_to = "Value") %>% 
  pivot_wider( names_from = "specification", values_from = "Value") %>% 
  mutate(`ROC Smoothed` = ifelse(`ROC Smoothed` == "smoothed", "Yes", "No"), 
         model = "Base Specification") %>% 
  rename(Escalation = escalation, 
         Quad = quad, 
         Goldstein = goldstein, 
         Avg = avg)

auc %>%
  knitr::kable(digits = 2, booktabs = TRUE, caption = "Comparison of smoothed and not smoothed AUC-ROC values", align = c(rep("l", 2), rep("c", 6))) %>% 
  # column_spec(c(3), bold = TRUE) %>%
  kableExtra::collapse_rows(1:2, row_group_label_position = 'stack')

@@

<!-- - Improperly tuning of models -->
<!--   - RM: In a twitter DM, I asked Blair to clarify -- "[I] was wondering how you all came to your tuning procedure for the RF model hyper-parameters?" He responded "mostly trial and error, honestly. ... Trial and error was on early data. Forecasts were for much later data" -->
<!-- - Using a regression framework in a classification problem -->
<!-- - smoothing of AUC/ROC scores -->
<!--     + AB: smoothing or rounding? I think rounding is ok.  -->
<!-- - Treating the output of a RF regression model as Pr() -- (RM: I guess this is akin to a Linear Probability Model, but I'm not sure if this is possible in a RF framework -->
<!--     + It's hacky but it works. The resulting forecasts are within the 0-1 interval. -->

@@

<!-- 4. Design choices -->

<!-- Rick, I think the points below are more subjective choices. The first two points above are IMO potentially objectively incorrect technical errors that could undermine the original Blair and Sambanis findings. The 3rd point about the RF models is somewhere between objective error and subjective choice. I'm not sure. But the points below I would say are subjective choices that could alter their findings and that I think are justifiable criticisms, but I think it'll be easier for someone to push back on this if they wanted to.  -->

<!-- - Their train/test approach  -->
<!--   - 5-year forecasts for sep plots and AUC scores) -->
<!-- - Smoothing of AUC/ROC scores -->
<!-- - Lack of yearly test forecasts in favor of a single 5-year test forecast -->
<!-- - The lack of procedures to account for rare events in an RF model -->

<!-- To-do -->

<!-- 1. Full replication -->
<!--     - Can we get their results -->
<!--     - RM: maybe I can go through their code and bring it "up-to-date" i.e. make it tidy...  -->
<!--         + AB: that sounds like a lot of work -->
<!--     - Figure out how they built table 4 (It looks like this is done in STATA) -->
<!--     - Check to see if it matter that they are using regression for a classification problem -->
<!--         + AB: i tried switching the models as they have them just from regression to classification and it breaks because the data samples end up being all 0's.  -->
<!-- 2. Address tuning  -->
<!--     - Does properly tuning the CAMEO model improve its performance? -->
<!-- 3. Address AUC/ROC smoothing...  -->
<!--     - How drastic is the change; does it change their conclusions?  -->
<!-- 4. Run a yearly test forecast -- as is the current standard in TSCS forecasts (Is it the current standard?)  -->
<!-- 5.  -->
